Ideological Positioning and Racial Media Framing in the Opioid Epidemic Influences Policy Outlooks

### 2024-02-20 paper/paper.qmd
Final commits to paper.pdf and paper.qmd, Feb 19 @ 7:53pm.

---
title: "Ideological Positioning and Racial Media Framing in the Opioid Epidemic Influences Policy Outlooks"
subtitle: "A Replication Study on Policy Outlooks among White Americans in the Context of Drug Abuse Media Narratives"
author: Bella MacLean
thanks: "*Code and data are available at: https://github.com/Bellamaclean7/Ideological-Influences-of-Opioid-Addiction-Media-Frames; Replication on Social Science Reproduction platform available at: https://doi.org/10.48152/ssrp-t3vw-5z41"
date: today
date-format: long
abstract: "This paper examines the interplay between political ideology, racial media framing, and public support for opioid abuse treatment policies in the United States, replicating and extending Tanika Raychaudhuri, Tali Mendelberg, and Anne McDonough's investigation into whether sympathy for drug users is racially selective among White Americans. Drawing on the original survey data, the paper explores how sympathetic and unsympathetic media portrayals of opioid abuse influence support for treatment policies across conservative and liberal ideologies. Findings reveal that there is an interaction effect between political ideology and media framing accross the ideoloigcal spectrum."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

This shift in the racial framing of drug use and policy within media and societal discourse is a direct result of White Americans and the the opioid epidemic. As a result, drug abuse has been met with a more sympathetic, medicalized response, highlighting a stark contrast in how drug use has historically been conceptualized and addressed throughout the country. A 2022 paper by Tanika Raychaudhuri, Tali Mendelberg, and Anne McDonough in The Journal of Politics
Volume 85, Number 1 (January 2023), titled "The Political Effects of Opioid Addiction Frames" delves into the impact of media coverage on public support for drug treatment policies, focusing on how the racial framing and valence of opioid abuse influence White Americans' policy views. Unlike previous drug epidemics portrayed negatively and associated with non-White users, the opioid crisis is often framed sympathetically and as predominantly affecting White Americans. This sympathetic portrayal is thought to contribute to the broad public support for treating opioid abuse as a public health issue rather than a criminal one [@Raychaudhuri2022].

@@

# Data {#sec-data}
## Source
The paper used for replication is from the The Journal of Politics Volume 85, Number 1
(January 2023) which examines whether sympathy for drug users is racially selective among White Americans [@Raychaudhuri2022]. This Reproduction seeks to utilize one of the studies major findings and then asses whether ideological tendencies influences the response to sympathetic media frames about opioid abuse, potentially conditioning the effect of racial selectivity in sympathy. This could help understand if conservative and liberal individuals are equally affected by the racial framing of opioid abuse stories.

## Methodology
This paper will replicate the survey data that was originally collected for the (2022) paper by Raychaudhuri, Mendelberg, and McDonough. The study constructed a hypothetical newspaper story about opioid abuse, varying the race of the users (Black or White) and the valence of the frame (sympathetic or unsympathetic). The focus on Black and White radicalized groups was guided by the White “face” of the opioid epidemic in contrast to the Black “face” of other drug epidemics [@Raychaudhuri2022]. The authors then conducted two randomized between-subjects survey experiments on two separate samples. The main sample consists of 1,517 White American adults from the NORC Amerispeak Panel, a national probability based sample. The other sample was from the survey firm Dynata. The authors describe the process as follows: "Each participant was randomly assigned to one of four treatments or a no-story control. In the treatments, respondents read a hypothetical news story about drug use, varying the race of users (White or Black) and the valence of the frame (sympathetic or unsympathetic)" [@Raychaudhuri2022].

# Literature {#sec-literature}
Much of the studies literature review draws on Michelle Alexander's 2010 book that focuses on mass incarceration in the age of colourblindness [@Alexander2010]. The paper also draw upon Michael Collins 2019 Article Titled “No Gentler War on Drugs: How Race Made the Opioid Crisis" [@Collins2019]. These peices of literature aid in theorizing that historically, drug abuse focuses on the racial impact of the “War on Drugs,” a punitive policy response to crack cocaine. Unlike the opioid crisis, the crack cocaine crisis led to the incarceration of millions, most of whom were non-White Americans [@Raychaudhuri2022]. Media coverage of cocaine often depicted Black users in an unsympathetic light, linking their drug use to involvement in other crimes. Julie Netherland and Helena Hansen's article titled "White opioids: Pharmaceutical race and the war on drugs that wasn’t" aids in proving that by contrast to the crack epidemic, news coverage now frames opioid abuse largely as a crisis affecting White Americans and when doing so it also tends to use a positive valence. By contrast, the Black “face” often carries negative overtones [@Raychaudhuri2022]. Overall, the literature aids in building the study's foundational argument that the racialization of drug epidemics can not merely be looked at as a matter of public health crises but also as an issue deeply intertwined with socio-political narratives and racial prejudices.

# Features
The original survey data assessed participants on 45 variables, being representative of geographical location, education level, polictical affiliation, ideology, and racial attitudes and affiliations. The main sample consists of 1,517 White American adults from the NORC Amerispeak Panel, a national probability based sample. The other sample was from the survey firm Dynata. The original paper includes three data sets; data from pre-test (Dynata panel), data from main experiment (NORC Amerispeak Panel), and data for the state-level opioid mortality analysis (CDC). However, this reproduction is focused solely on the data from the main experiement and includes variables only found in the main experiment data set. This reproduction omits data for the state-level opioid mortality analysis (CDC) and the data from the pre-test (Dynata panel). This reproduction also changed the name of some of the main experiments variables in the data set for better clarity. The data set was then cleaned to include only relevant variables applicable to this reproduction. @tbl-cleaned_data shows a sample of the cleaned data set. @tbl-variable_table discusses the measurement of each variable used in the replication.

@@

@@

# Results  {#sec-results}
## Policy Support by Media Frame
@fig-policy_vs_media_frames is a replication of Raychaudhuri, Mendelberg, and McDonough's "Figure A5: Mean outcome values across experimental conditions (raw values, 83% CIs)", found in the Appendix in A.5. This replication is focused on only the first graph of the quadrant (A.5.A). The purpose of this graph is to visualize the mean policy preference for each media frame, essentially determining the proportion of respondents in each category whom support treatment. This replication attempts to simplify the original by demonstrating the findings in graph that is more easily digestible to the average audience.

>>>>

---
title: "Ideological Positioning and Racial Media Framing in the Opioid Epidemic Influences Policy Outlooks"
subtitle: "A Replication Study on Policy Outlooks among White Americans in the Context of Drug Abuse Media Narratives"
author: Bella MacLean
thanks: "*Code and data are available at: https://github.com/Bellamaclean7/Ideological-Influences-of-Opioid-Addiction-Media-Frames; Replication on Social Science Reproduction platform available at: https://doi.org/10.48152/ssrp-t3vw-5z41"
date: today
date-format: long
abstract: "This paper examines the interplay between political ideology, racial media framing, and public support for opioid abuse treatment policies in the United States. It replicates and extends the investigation by Tanika Raychaudhuri, Tali Mendelberg, and Anne McDonough into whether sympathy for drug users among White Americans is racially selective. Utilizing the original survey data, this study explores how sympathetic and unsympathetic media portrayals of opioid abuse influence support for treatment policies across conservative and liberal ideologies. The findings reveal an interaction effect between political ideology and media framing across the ideological spectrum."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

This shift in the racial framing of drug use and policy within media and societal discourse is a direct result of White Americans and the opioid epidemic. As a result, drug abuse has been met with a more sympathetic, medicalized response, highlighting a stark contrast in how drug use has historically been conceptualized and addressed throughout the country. A 2022 paper by Tanika Raychaudhuri, Tali Mendelberg, and Anne McDonough in The Journal of Politics
Volume 85, Number 1 (January 2023), titled "The Political Effects of Opioid Addiction Frames" delves into the impact of media coverage on public support for drug treatment policies, focusing on how the racial framing and valence of opioid abuse influence White Americans' policy views. Unlike previous drug epidemics portrayed negatively and associated with non-White users, the opioid crisis is often framed sympathetically and as predominantly affecting White Americans. This sympathetic portrayal is thought to contribute to the broad public support for treating opioid abuse as a public health issue rather than a criminal one [@Raychaudhuri2022].

@@

# Data {#sec-data}
## Source
The paper used for replication is from The Journal of Politics Volume 85, Number 1
(January 2023) which examines whether sympathy for drug users is racially selective among White Americans [@Raychaudhuri2022]. This Reproduction seeks to utilize one of the study's major findings and then assess whether ideological tendencies influences the response to sympathetic media frames about opioid abuse, potentially conditioning the effect of racial selectivity in sympathy. This could help understand if conservative and liberal individuals are equally affected by the racial framing of opioid abuse stories.

## Methodology
This paper will replicate the survey data that was originally collected for the 2022 paper by Raychaudhuri, Mendelberg, and McDonough. The study constructed a hypothetical newspaper story about opioid abuse, varying the race of the users (Black or White) and the valence of the frame (sympathetic or unsympathetic). The focus on Black and White racialized groups was guided by the White 'face' of the opioid epidemic in contrast to the Black 'face' of other drug epidemics [@Raychaudhuri2022]. The authors then conducted two randomized between-subjects survey experiments on two separate samples. The main sample consists of 1,517 White American adults from the NORC Amerispeak Panel, a national probability-based sample. The other sample was from the survey firm Dynata. The authors describe the process as follows: 'Each participant was randomly assigned to one of four treatments or a no-story control. In the treatments, respondents read a hypothetical news story about drug use, varying the race of users (White or Black) and the valence of the frame (sympathetic or unsympathetic)' [@Raychaudhuri2022].

@@

## Features
The original survey data assessed participants on 45 variables, being representative of geographical location, education level, political affiliation, ideology, and racial attitudes and affiliations. The main sample consists of 1,517 White American adults from the NORC Amerispeak Panel, a national probability based sample. The other sample was from the survey firm Dynata. The original paper includes three data sets; data from pre-test (Dynata panel), data from main experiment (NORC Amerispeak Panel), and data for the state-level opioid mortality analysis (CDC). However, this reproduction is focused solely on the data from the main experiment and includes variables only found in the main experiment data set. This reproduction omits data for the state-level opioid mortality analysis (CDC) and the data from the pre-test (Dynata panel). This reproduction also changed the name of some of the main experiments variables in the data set for better clarity. The data set was then cleaned to include only relevant variables applicable to this reproduction. @tbl-cleaned_data shows a sample of the cleaned data set. @tbl-variable_table discusses the measurement of each variable used in the replication.

# Literature {#sec-literature}
Much of the study's literature review draws on Michelle Alexander's 2010 book that focuses on mass incarceration in the age of colourblindness [@Alexander2010]. The paper also draws upon Michael Collins 2019 Article Titled “No Gentler War on Drugs: How Race Made the Opioid Crisis" [@Collins2019]. These pieces of literature aid in theorizing that historically, drug abuse focuses on the racial impact of the “War on Drugs,” a punitive policy response to crack cocaine. Unlike the opioid crisis, the crack cocaine crisis led to the incarceration of millions, most of whom were non-White Americans [@Raychaudhuri2022]. Media coverage of cocaine often depicted Black users in an unsympathetic light, linking their drug use to involvement in other crimes. Julie Netherland and Helena Hansen's article titled "White opioids: Pharmaceutical race and the war on drugs that wasn’t" aids in proving that by contrast to the crack epidemic, news coverage now frames opioid abuse largely as a crisis affecting White Americans and when doing so it also tends to use a positive valence. By contrast, the Black “face” often carries negative overtones [@Raychaudhuri2022]. Overall, the literature aids in building the study's foundational argument that the racialization of drug epidemics can not merely be looked at as a matter of public health crises but also as an issue deeply intertwined with socio-political narratives and racial prejudices.

@@

# Results  {#sec-results}
## Policy Support by Media Frame
@fig-policy_vs_media_frames is a replication of Raychaudhuri, Mendelberg, and McDonough's "Figure A5: Mean outcome values across experimental conditions", found in the Appendix in A.5. This replication is focused on only the first graph of the quadrant (A.5.A). The purpose of this graph is to visualize the mean policy preference for each media frame, essentially determining the proportion of respondents in each category whom support treatment. This replication attempts to simplify the original by demonstrating the findings in graph that is more easily digestible to the average audience.
